Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-11-08 acks.tex
added an acknowledgement for Danny's COPE grant
final popl17 submission

Thanks to Ed Morehouse for counsel on mechanization and Barendrecht's
Convention, and to Vincent Zeng for \textit{pro bono} artistic services.

>>>>

% !TEX root = hazelnut-popl17.tex

The authors would like to thank the anonymous referees at POPL and TFP 2016  for useful feedback on earlier drafts of this paper; Ed Morehouse and Carlo Angiuli for counsel on mechanization and Barendrecht's
Convention; Vincent Zeng for \textit{pro bono} artistic services; and YoungSeok Yoon for reanalyzing the data from \cite{6883030}. 
%This is Danny Dig's COPE grant. He asked me to add the grant.
This work was partially funded through the NSF grant \#CCF-1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140. The views and conclusions contained herein are those of the authors and should not be interpreted as necessarily representing the official policies or endorsements, either expressed or implied, of the NSF, AFRL, DARPA, the NSA or the U.S. Government.

### 2016-11-08 example.tex
revised abstract and intro
final popl17 submission

% !TEX root = hazelnut-popl17.tex
The reader is encouraged to follow along with the examples that we will
discuss in this section using the implementation.

\subsection{Example 1: Constructing the Increment Function}

Figure~\ref{fig:first-example} shows an edit sequence that constructs the
increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty
hole via the indicated sequence of {actions}. We will introduce Hazelnut's
formal syntax and define the referenced rules in
Sec. \ref{sec:hazel}. First, let us build some high-level intuitions.

The edit state in Hazelnut is a {Z-expression}, $\zexp$. Every Z-expression
has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the
{cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively. For
example, on Line 1, the empty expression hole, $\hhole{}$, is under the
cursor.

Actions act relative to the cursor. The first action that we perform is
$\aConstruct{\flam{x}}$, which instantiates the hole with a lambda
abstraction binding the variable $x$. This results in the Z-expression on
Line 2, consisting of a lambda abstraction with an arrow type
ascription. The cursor is placed on the argument type hole. Type holes are
typeset like empty expression holes for visual consistency.

The actions on Lines 2-5 complete the type ascription. In particular, the
$\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
composition of the $\aMove{\dParent}$ and $\aMove{\dChildn{2}}$ actions
move the cursor to the next hole.

\begin{figure}[t!]
\begin{center}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} &
\aConstruct{\flam{x}} &
\text{(\ref{r:conelamhole})}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
\aMove{\dParent} &
\text{({\ref{rule:move-parent-arr-left}})}
\\ 4 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\hhole{}}} &
\aMove{\dChildn{2}} &
\text{({\ref{rule:move-arr-c2}})}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 6 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-arr-right})}
\\ 7 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-asc-right})}
\\ 8 &
\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} &
\aMove{\dChildn{1}} &
\text{(\ref{r:movefirstchild-asc})}
\\ 9 &
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
\aMove{\dChildn{1}} &
\text{(\ref{r:movefirstchild-lam})}
\\ 10 &
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fvar{x}} &
\text{(\ref{r:conevar})}
\\ 11 &
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 12 &
\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fnumlit{1}} &
\text{(\ref{r:conenumnum})}
\\ 13 &
\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing an increment function in Hazelnut.}
\label{fig:first-example}
\end{figure}

The actions on Lines 6-9 move the cursor to the function body. For
simplicity, Hazelnut defines only a small collection of primitive movement
actions. In practice, an editor would also define compound movement
actions, e.g. an action that moves the cursor directly to the next
remaining hole, in terms of these primitive movement actions.

Lines 10-12 complete the function body by first constructing the variable
$x$, then constructing the plus form, and finally constructing the number
$\hnum{1}$. Notice that we did not need to construct the function body in
an ``outside-in'' manner, i.e. we constructed $x$ before constructing the
outer plus form. Luckily, the transient function bodies, $x$ and $x +
\hhole{}$, could be checked against type $\tnum$ (as we will detail in
Sec. \ref{sec:holes}.)

@@

\begin{figure}[t!]
\begin{center}
\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
14 &
\zwsel{\hhole{}} &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}&
\text{(\ref{r:conevar})}
\\ 15 &
\zwsel{incr} &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 16 &
incr(\zwsel{\hhole{}}) &
\aConstruct{\fvar{incr}} &
\text{(\ref{r:conevar2})}
\\ 17 &
incr(\hhole{\zwsel{incr}}) &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 18 &
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} &
\aConstruct{\fnumlit{3}} &
\text{(\ref{r:conenumnum})}
\\ 19 &
incr(\hhole{incr(\zwsel{\hnum{3}})}) &
\aMove{\dParent}&
\text{(\ref{r:moveparent-ap2})}
\\ 20 &
incr(\hhole{\zwsel{incr(\hnum{3})}}) &
\aMove{\dParent} &
\text{(\ref{r:moveparent-hole})}
\\ 21 &
incr(\zwsel{\hhole{incr(\hnum{3})}})&
\aFinish &
\text{(\ref{r:finishana})}
\\ 22 &
incr(\zwsel{incr(\hnum{3})}) &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Appyling the increment function.}
\label{fig:second-example}
\end{figure}

@@

We begin on Line 14 by constructing the variable $incr$. Line 15 then
constructs the application form with $incr$ in function position, leaving
the cursor on a hole in the argument position. Notice that we did not need
to construct the outer application form before identifying the function
being applied. Again, the transient edit state happens to be well-typed, so
we needed to make no special allowances for this order of actions.

We now need to apply $incr$ again, so we perform the same action on Line 16
as we did on Line 14, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
structure editor, performing such an action would result in the following
edit state:
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed (after \emph{cursor erasure}): the argument of
$incr$ must be of type $\tnum$ but here it is of type
$\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.

The programmer could alternatively have performed the $\aConstruct{\fap}$
action on Line 16. This would result in the following edit state, which is
well-typed according to the static semantics that we will define in the
next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended
function before constructing the function application form. This stands in
contrast to Lines 14-15.

>>>>

% !TEX root = hazelnut-popl17.tex
\subsection{Example 1: Constructing the Increment Function}

Figure~\ref{fig:first-example} shows an edit sequence that constructs the
increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty
hole via the indicated sequence of {actions}. We will introduce Hazelnut's
formal syntax and define the referenced rules in
Sec. \ref{sec:hazel}.\footnote{For concision, the column labeled \textbf{Rule} in Figures \ref{fig:first-example} and \ref{fig:second-example} 
indicates only the relevant \emph{non-zipper} rule (see Sec. \ref{sec:zipper-cases}.) The reader is encouraged to follow along with these examples using the reference implementation. The derivations for these examples are also available in the Agda mechanization.} First, let us build some high-level intuitions.

The edit state in Hazelnut is a {Z-expression}, $\zexp$. Every Z-expression
has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the
{cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively.\footnote{The reference implementation omits the triangles, while the Agda mechanization necessarily omits the colors.} For
example, on Line 1, the empty expression hole, $\hhole{}$, is under the
cursor.

Actions act relative to the cursor. The first action, on Line 1, is
$\aConstruct{\flam{x}}$. This results in the Z-expression on
Line 2, consisting of a lambda abstraction with argument $x$ under an arrow type
ascription. The cursor is placed (arbitrarily) on the argument type hole.

The actions on Lines 2-5 complete the type ascription. In particular, the
$\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
the $\aMove{\dParent}$ and $\aMove{\dChildn{2}}$ action sequence 
moves the cursor to the next hole. In practice, an editor would also define compound movement
actions, e.g. an action that moves the cursor directly to the next
 hole, in terms of these primitive movement actions.

\begin{figure}[t!]
  \label{ex1}
\begin{center}
$\arraycolsep=2px
\begin{array}[p]{|c||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} &
\aConstruct{\flam{x}} &
\text{(\ref{r:conelamhole})}
\\ 2 &
{\hlam{x}{\hhole{}}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
{\hlam{x}{\hhole{}}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
\aMove{\dParent} &
\text{({\ref{rule:move-parent-arr-left}})}
\\ 4 &
{\hlam{x}{\hhole{}}} : \zwsel{\tarr{\tnum{}}{\hhole{}}} &
\aMove{\dChildn{2}} &
\text{({\ref{rule:move-arr-c2}})}
\\ 5 &
{\hlam{x}{\hhole{}}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 6 &
{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-arr-right})}
\\ 7 &
{\hlam{x}{\hhole{}}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-asc-right})}
\\ 8 &
\zwsel{{\hlam{x}{\hhole{}}} : \tarr{\tnum}{\tnum}} &
\aMove{\dChildn{1}} &
\text{(\ref{r:movefirstchild-asc})}
\\ 9 &
{\zwsel{\hlam{x}{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
\aMove{\dChildn{1}} &
\text{(\ref{r:movefirstchild-lam})}
\\ 10 &
{\hlam{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fvar{x}} &
\text{(\ref{r:conevar})}
\\ 11 &
{\hlam{x}{\zwsel{{x}}}} : \tarr{\tnum{}}{\tnum{}} &
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 12 &
{\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fnumlit{1}} &
\text{(\ref{r:conenumnum})}
\\ 13 &
{\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}}} : \tarr{\tnum}{\tnum} &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing the increment function in Hazelnut.}
\label{fig:first-example}
\end{figure}

After moving to the function body, Lines 10-12 complete it by first constructing the variable
$x$, then constructing the plus form, and finally constructing the number literal 
$\hnum{1}$. Notice that we did not need to construct the function body in
an ``outside-in'' manner, i.e. we constructed $x$ before constructing the
outer plus form inside which $x$ ultimately appears. The transient function bodies, $x$ and $\hadd{x}{\hhole{}}$, can be checked against the given return type, $\tnum$ (as we will detail in
Sec. \ref{sec:holes}.)

@@

\begin{figure}[t!]
  \label{ex2}
\begin{center}
\colorbox{light-gray}{\hspace{54px} now assume $incr : \tarrnp{\tnum}{\tnum}$ \hspace{53px}}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
14 &
\zwsel{\hhole{}} &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}&
\text{(\ref{r:conevar})}
\\ 15 &
\zwsel{incr} &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 16 &
incr(\zwsel{\hhole{}}) &
\aConstruct{\fvar{incr}} &
\text{(\ref{r:conevar2})}
\\ 17 &
incr(\hhole{\zwsel{incr}}) &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 18 &
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} &
\aConstruct{\fnumlit{3}} &
\text{(\ref{r:conenumnum})}
\\ 19 &
incr(\hhole{incr(\zwsel{\hnum{3}})}) &
\aMove{\dParent}&
\text{(\ref{r:moveparent-ap2})}
\\ 20 &
incr(\hhole{\zwsel{incr(\hnum{3})}}) &
\aMove{\dParent} &
\text{(\ref{r:moveparent-hole})}
\\ 21 &
incr(\zwsel{\hhole{incr(\hnum{3})}})&
\aFinish &
\text{(\ref{r:finishana})}
\\ 22 &
incr(\zwsel{incr(\hnum{3})}) &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Applying the increment function.}
\label{fig:second-example}
\end{figure}

@@

We begin on Line 14 by constructing the variable $incr$. Line 15 then
constructs the application form with $incr$ in function position, leaving
the cursor on a hole in the argument position. Notice again that we did not construct the outer application form before identifying the function
being applied. %Again, the transient H-expression happens to be well-typed.

We now need to apply $incr$ again, so we perform the same action on Line 16
as we did on Line 14, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
structure editor, performing such an action would result in the following
edit state:
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed (after \emph{cursor erasure}): the argument of
$incr$ must be of type $\tnum$ but here it is of type
$\tarr{\tnum}{\tnum}$. Hazelnut does not allow such an edit state to arise.

We could alternatively have performed the $\aConstruct{\fap}$
action on Line 16. This would result in the following edit state, which is
well-typed according to the static semantics that we will define in the
next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended
function, $incr$, before constructing the function application form. This stands in
contrast to Lines 14-15.

### 2016-11-08 intro.tex
final popl17 submission

First, it requires that programmers learn the subtleties of the textual
syntax (e.g. operator precedence.) This is particularly 
challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258,stefik2013empirical}, and even experienced programmers frequently make mistakes \cite{7548903,jones2006developer}.

Second, the fact that many sequences of characters do not correspond to a meaningful
program complicates the design of program editors and other interactive programming tools. 
%In particular, program
%editors must contend with meaningless text on a regular basis. 
In a dataset  
gathered by Yoon and Myers consisting of 1460 hours of Eclipse edit logs  \cite{6883030}, 44.2\% of
edit states were syntactically malformed. Some additional
percentage of edit states were well-formed but ill-typed (the dataset was
not complete enough to determine the exact percentage.) Collectively, we refer to these edit states as \emph{meaningless edit states}, because they are not given static or dynamic meaning by the language definition. 
%Some of
%these meaningless edit states arose because the programmer was in the midst
%of a sequence of edits that left the edit state transiently meaningless,
%while others arose because the programmer made a logical mistake. 
As a consequence, it is difficult
to design useful language-aware editor services, e.g. syntax
highlighting~\cite{sarkar2015impact}, type-aware code
completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324},
and refactoring support \cite{mens2004survey}. Editors must either disable
these editor services when they encounter meaningless edit states or
deploy \emph{ad hoc} heuristics, e.g. by using whitespace to guess the programmer's intent \cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.

These complications have motivated a long line of research
into \emph{structure editors}, i.e. program editors where every edit state
corresponds to a program structure \cite{teitelbaum_cornell_1981}.\footnote{Structure editors are also variously known as \emph{structured editors}, \emph{structural editors}, \emph{syntax-directed editors} and \emph{projectional editors} (esp. when the user interface presents rich views of the program structure.)}

@@

% Scratch does not impose any static requirements , i.e. every well-formed edit state is trivially statically meaningful. 
Researchers have also 
 designed  syntactic structure editors for more sophisticated languages with non-trivial  
static type systems. Contemporary examples include \texttt{mbeddr}, a structure editor for a C-like
language \cite{voelter_mbeddr:_2012}, TouchDevelop, a structure editor for an
object-oriented language \cite{tillmann_touchdevelop:_2011}, and Lamdu, a structure 
editor for a functional language similar to Haskell \cite{lamdu}. Each of
these editors presents an innovative user interface, but the non-trivial
type and binding structure of the underlying language complicates its
design. The reason is that syntactic structure editors do not assign static meaning to every edit state -- they guarantee only that every edit state corresponds to  a 
syntactically well-formed tree. These editors must also either disable editor services that require an understanding of the semantics of the program or deploy \emph{ad hoc} heuristics when they
encounter these meaningless edit states.

>>>>

First, it requires that programmers learn the subtleties of the textual
syntax (e.g. operator precedence.) This is particularly 
challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258,stefik2013empirical,Marceau:2011:VGT:2016911.2016921}, and even experienced programmers frequently make mistakes \cite{7548903,jones2006developer,Marceau:2011:VGT:2016911.2016921}.

Second, many sequences of characters do not correspond to meaningful
programs. This complicates the design of program editors and other interactive programming tools. 
%In particular, program
%editors must contend with meaningless text on a regular basis. 
In a dataset  
gathered by Yoon and Myers consisting of 1460 hours of Eclipse edit logs  \cite{6883030}, 44.2\% of
edit states were syntactically malformed. Some additional
percentage of edit states were well-formed but ill-typed (the dataset was
not complete enough to determine the exact percentage.) Collectively, we refer to these edit states as \emph{meaningless edit states}, because they are not given static or dynamic meaning by the language definition. 
%Some of
%these meaningless edit states arose because the programmer was in the midst
%of a sequence of edits that left the edit state transiently meaningless,
%while others arose because the programmer made a logical mistake. 
As a consequence, it is difficult
to design useful language-aware editor services, e.g. syntax
highlighting~\cite{sarkar2015impact}, type-aware code
completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324},
and refactoring services \cite{mens2004survey}. Editors must either disable
these editor services when they encounter meaningless edit states or
deploy \emph{ad hoc} heuristics, e.g. by using whitespace to guess the intent \cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.

These complications have motivated a long line of research
into \emph{structure editors}, i.e. program editors where every edit state
corresponds to a program structure \cite{teitelbaum_cornell_1981}.\footnote{Structure editors are also variously known as \emph{structured editors}, \emph{structural editors}, \emph{syntax-directed editors} and \emph{projectional editors}.}

@@

% Scratch does not impose any static requirements , i.e. every well-formed edit state is trivially statically meaningful. 
Researchers have also 
 designed  syntactic structure editors for more sophisticated languages with non-trivial  
static type systems. Contemporary examples include \texttt{mbeddr}, a structure editor for a C-like
language \cite{voelter_mbeddr:_2012}, TouchDevelop, a structure editor for an
object-oriented language \cite{tillmann_touchdevelop:_2011}, and Lamdu, a structure 
editor for a functional language similar to Haskell \cite{lamdu}. Each of
these editors presents an innovative user interface, but the non-trivial
type and binding structure of the underlying language complicates its
design. The problem is that syntactic structure editors do not assign static meaning to every edit state -- they guarantee only that every edit state corresponds to  a 
syntactically well-formed tree. These editors must also either selectively disable editor services that require an understanding of the semantics of the program being written, or deploy \emph{ad hoc} heuristics.
